{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-07 thesis/tm.tex
llamas and studence
et cetera
write intro/motivation for htm

The example TSX program from \Cref{fig:htm-example} (\cref{sec:overview-tm})
is reproduced here for the reader's convenience.

@@

\begin{figure}[h]
	\begin{center}
		\begin{tabular}{ll}
			1 & \texttt{\flow{if} ((status = \call{\_xbegin}()) == \const{\_XBEGIN\_STARTED}) \{} \\
			%2 & \texttt{~~~~x++;} \\
			$2a$ & \texttt{~~~~temp <- x;} \\
			$2b$ & \texttt{~~~~temp <- temp + 1;} \\
			$2c$ & \texttt{~~~~x <- temp;} \\
			3 & \texttt{~~~~\call{\_xend}();} \\
			4 & \texttt{\} \flow{else} \{} \\
			5 & \texttt{~~~~\call{mutex\_lock}(\&m);} \\
			$6a$ & \texttt{~~~~temp <- x;} \\
			$6b$ & \texttt{~~~~temp <- temp + 1;} \\
			$6c$ & \texttt{~~~~x <- temp;} \\
			7 & \texttt{~~~~\call{mutex\_unlock}(\&m);} \\
			8 & \texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption{\Cref{fig:htm-example-reproduced} with non-atomic operations shown as pseudo-assembly.}
	\label{fig:htm-example-assemblified}
\end{figure}

@@

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{ll}
			\hilight{lavender}{\bf Thread 1} &
			\hilight{seafoam}{\bf Thread 2} \\
			\hline
			\texttt{status = \call{\_xbegin}()} & \\
			\texttt{temp <- x;} & \\
			\texttt{temp <- temp + 1;} & \\
			& \texttt{status = \call{\_xbegin}()} \\
			& \texttt{x++;} \\
			& \texttt{\call{\_xend}();} \\
			\texttt{x <- temp;} & \\
			\texttt{\call{\_xend}();} & \\
		\end{tabular}
	\end{center}
	\caption[Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-reproduced}.]
		{Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-reproduced}.
		The effects of $\tiat{2a-2b}$ are never globally visible, as its transasction must abort,
		and moreover $\tiat{2c-3}$ will not execute at all.}
	\label{fig:htm-example-interleaving}
\end{figure}

@@

\begin{lemma}[Equivalence of Aborts]
	\label{lem:equiv}
	For any interleaving prefix
	\[
	\begin{tabular}{c}
		$\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b},$\\
		$\tjjat{\gamma_1}\dots\tjjat{\gamma_j},$ \\
		$\tkkat{\kappa_1}\dots\tkkat{\kappa_k},$ \\
		$\tiiat{\beta_{b+1}}$ % $\dots\tiiat{\beta_{n-1}}$ -- excluded bc might abort
	\end{tabular}
	\]
	with $b<n$, $j \ne i$, $k \ne i$, etc., either:
	\begin{enumerate}
		\item $\tiiat{\alpha},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k},\tiiat{\phi_1}\dots$
			(conflicting case), or
		\item $\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b}\dots\tiiat{\beta_n},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
			(independent case)
	\end{enumerate}
	exists and is observationally equivalent.
\end{lemma}

@@

\begin{itemize}
	\item Unit tests and microbenchmarks
	\begin{itemize}
		\item {\tt htm1}: The bug from \Cref{fig:htm-example,fig:htm-example-reproduced}.
			%(assertion failure on {\tt count}).
		\item {\tt htm2}: The fixed version as in \Cref{fig:htm-fixed}.
		\item {\tt counter}: Microbenchmark version of {\tt htm2} which replaces the complex locking failure path with an atomic {\tt xadd}, from \cite{tm-benchmark-cmu}.
		\item {\tt swap}: Microbenchmark that swaps values in an array, from \cite{tm-benchmark-cmu}.
		\item {\tt swapbug}: {\tt swap} modified to introduce circular locking in the failure path. %(deadlock).
			% TODO: make sure figure number reference is right
		\item {\tt fig63}: Generalized version of \Cref{fig:retry-set-example},
			contrived to induce as much reduction as possible from the retry sets optimization.
	\end{itemize}
	\item Data structure tests
	\begin{itemize}
		\item {\tt avl\_insert}: AVL tree concurrent insertion test \cite{tm-benchmark-cmu}.
		\item {\tt avl\_fixed}: {\tt avl\_insert} with the AVL bug fixed (spoilers!!).
		\item {\tt map\_basic}: Separate-chaining hashmap concurrent insertion test \cite{tm-benchmark-cmu}.
		\item {\tt map\_basicer}: {\tt map\_basic} modified with a larger initial size to skip the resizing step.
		\item {\tt avl\_mutex}: {\tt avl\_fixed} with transactional sections simplified by abstraction into a mutex.
		\item {\tt map\_mutex}: {\tt map\_basic} simplified likewise.
		\item {\tt map\_basicer\_mutex}: {\tt map\_basicer} simplified likewise.
	\end{itemize}
	\item Lock abstraction tests
	\begin{itemize}
		\item {\tt lock}(): Checks that multiple threads using a transactional lock
			cannot access the critical section simultaneously.
			% htm_spinlock and htm_mutex
		\item {\tt lock\_fast}(): Checks that a transactional lock's fast path will not suffer conflict aborts
			if its client threads' critical sections are independent.
			% htm_spinlock_fastpath and htm_mutex_fastpath
	\end{itemize}
		These are each parameterized over implementations {\tt spinlock} (from \cite{spinlock-rtm-github}),
		{\tt spin\_\allowbreak{}fixed} (spoilers!!),
		and {\tt mutex} (replaces the spinlock with a Landslide-annotated P2 mutex to reduce state space size).
\end{itemize}

>>>>

% or "Contribution"
\revision{{\bf Motivation.}
For the most part,
existing stateless model checkers
model interrupt-driven thread switching as the only source of nondeterminism in a program's execution.
In other words, assuming a fixed test input,%
\footnote{Input nondeterminism is another research field,
addressed by symbolic execution \cite{symbolic}; see \cref{related:symbolic}.}
the only way multiple behaviours can arise is by scheduling different threads at each preemption point.
Only recently have checkers emerged that extend this model by
incorporating other nondeterminism sources:
store buffer reordering arising from relaxed memory models
\cite{cdschecker,tsopso,rcmc}
and event-driven application programming models
\cite{r4,eventracer},
both of which are especially relevant in mobile programming.
Under TM, {\sf begin}'s return value is also nondeterministic,
and moreover may depend on execution events from the future in a parallel universe
(i.e., subsequent memory conflicts detected and reverted),
and no model checker yet exists which can model this.
In this chapter I will develop a concurrency model to incorporate abort nondeterminism
(\cref{sec:tm-design}),
including two equivalence proofs to help cope with the consequent state space explosion
(\cref{sec:tm-design-failure} and \cref{sec:tm-design-locks}),
extend Landslide to support testing TM programs under a variety of execution semantics
(\cref{sec:tm-implementation}),
and demonstrate its practicality
%\cref{sec:tm-eval}
both by finding new bugs
(\cref{sec:tm-eval-bugs})
and by verifying correctness
(\cref{sec:tm-verif})
in a variety of microbenchmarks and real-world HTM programs.}

\revisionminor{The following sections will make heavy reference
to the example TSX program from \Cref{fig:htm-example} (\cref{sec:overview-tm})
while defining the new concurrency model,
so I reproduce it here in \Cref{fig:htm-example-reproduced} for the reader's convenience.}

@@

\revision{
\begin{figure}[h]
	\begin{center}
		\begin{tabular}{ll}
			1 & \texttt{\flow{if} ((status = \call{\_xbegin}()) == \const{\_XBEGIN\_STARTED}) \{} \\
			%2 & \texttt{~~~~x++;} \\
			$2a$ & \texttt{~~~~temp <- x;} \\
			$2b$ & \texttt{~~~~temp <- temp + 1;} \\
			$2c$ & \texttt{~~~~x <- temp;} \\
			3 & \texttt{~~~~\call{\_xend}();} \\
			4 & \texttt{\} \flow{else} \{} \\
			5 & \texttt{~~~~\call{mutex\_lock}(\&m);} \\
			$6a$ & \texttt{~~~~temp <- x;} \\
			$6b$ & \texttt{~~~~temp <- temp + 1;} \\
			$6c$ & \texttt{~~~~x <- temp;} \\
			7 & \texttt{~~~~\call{mutex\_unlock}(\&m);} \\
			8 & \texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption{\Cref{fig:htm-example-reproduced} with non-atomic operations shown as pseudo-assembly.}
	\label{fig:htm-example-assemblified}
\end{figure}

@@

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{ll}
			\hilight{lavender}{\bf Thread 1} &
			\hilight{seafoam}{\bf Thread 2} \\
			\hline
			\texttt{status = \call{\_xbegin}()} & \\
			\texttt{temp <- x;} & \\
			\texttt{temp <- temp + 1;} & \\
			& \texttt{status = \call{\_xbegin}()} \\
			& \texttt{x++;} \\
			& \texttt{\call{\_xend}();} \\
			\texttt{x <- temp;} & \\
			\texttt{\call{\_xend}();} & \\
		\end{tabular}
	\end{center}
	\caption[Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-assemblified}.]
		{Impossible hypothetical interleaving within transactions of \Cref{fig:htm-example-assemblified}.
		The effects of $\tiat{2a-2b}$ are never globally visible, as its transasction must abort,
		and moreover $\tiat{2c-3}$ will not execute at all.}
	\label{fig:htm-example-interleaving}
\end{figure}
}

@@

\begin{lemma}[Equivalence of Aborts]
	\label{lem:equiv}
	For any interleaving prefix
	\[
	\begin{tabular}{c}
		$\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b},$\\
		$\tjjat{\gamma_1}\dots\tjjat{\gamma_j},$ \\
		$\tkkat{\kappa_1}\dots\tkkat{\kappa_k},$ \\
		$\tiiat{\beta_{b+1}}$ % $\dots\tiiat{\beta_{n-1}}$ -- excluded bc might abort
	\end{tabular}
	\]
	with $b<n$, $j \ne i$, $k \ne i$, et cetera, either:
	\begin{enumerate}
		\item $\tiiat{\alpha},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k},\tiiat{\phi_1}\dots$
			(conflicting case), or
		\item $\tiiat{\alpha},\tiiat{\beta_1}\dots\tiiat{\beta_b}\dots\tiiat{\beta_n},\tjjat{\gamma_1}\dots\tjjat{\gamma_j},\tkkat{\kappa_1}\dots\tkkat{\kappa_k}$
			(independent case)
	\end{enumerate}
	exists and is observationally equivalent.
\end{lemma}

@@

\begin{enumerate}
	\item Unit tests and microbenchmarks
	\begin{itemize}
		\item {\tt htm1}: The bug from \Cref{fig:htm-example,fig:htm-example-reproduced}.
			%(assertion failure on {\tt count}).
		\item {\tt htm2}: The fixed version as in \Cref{fig:htm-fixed}.
		\item {\tt counter}: Microbenchmark version of {\tt htm2} which replaces the complex locking failure path with an atomic {\tt xadd}, from \cite{tm-benchmark-cmu}.
		\item {\tt swap}: Microbenchmark that swaps values in an array, from \cite{tm-benchmark-cmu}.
		\item {\tt swapbug}: {\tt swap} modified to introduce circular locking in the failure path. %(deadlock).
			% TODO: make sure figure number reference is right
		\item {\tt fig63}: Generalized version of \Cref{fig:retry-set-example},
			contrived to induce as much reduction as possible from the retry sets optimization.
	\end{itemize}
	\item Data structure tests
	\begin{itemize}
		\item {\tt avl\_insert}: AVL tree concurrent insertion test \cite{tm-benchmark-cmu}.
		\item {\tt avl\_fixed}: {\tt avl\_insert} with the AVL bug fixed (spoilers!!).
		\item {\tt map\_basic}: Separate-chaining hashmap concurrent insertion test \cite{tm-benchmark-cmu}.
		\item {\tt map\_basicer}: {\tt map\_basic} modified with a larger initial size to skip the resizing step.
		\item {\tt avl\_mutex}: {\tt avl\_fixed} with transactional sections simplified by abstraction into a mutex.
		\item {\tt map\_mutex}: {\tt map\_basic} simplified likewise.
		\item {\tt map\_basicer\_mutex}: {\tt map\_basicer} simplified likewise.
	\end{itemize}
	\item Lock abstraction tests
	\begin{itemize}
		\item {\tt lock}(): Checks that multiple threads using a transactional lock
			cannot access the critical section simultaneously.
			% htm_spinlock and htm_mutex
		\item {\tt lock\_fast}(): Checks that a transactional lock's fast path will not suffer conflict aborts
			if its client threads' critical sections are independent.
			% htm_spinlock_fastpath and htm_mutex_fastpath
	\end{itemize}
		These are each parameterized over implementations {\tt spinlock} (from \cite{spinlock-rtm-github}),
		{\tt spin\_\allowbreak{}fixed} (spoilers!!),
		and {\tt mutex} (replaces the spinlock with a Landslide-annotated P2 mutex to reduce state space size).
\end{enumerate}
